---
title: "Simulating a Market with an Agent-Based Model"
author: Rayhan Ali

# --- Editorial title-slide fields (all optional; delete any you don't want) ---
kicker: "Projects"          # top-left label
issue: "Q3 2026"               # top-centre label
page-marker: "01"              # top-right label
prepared-for: "Public"         # right-hand metadata column
status: "Write-up"
author-label: "Author"         # heading for the author row
tagline: "Watching supply and demand build themselves out of buyers and sellers who each know only one price"

format:
  revealjs:
    theme: [default, custom.scss]
    template-partials:
      - title-slide.html
    width: 1280            # 16:9, PowerPoint-style canvas
    height: 720
    margin: 0.04
    fontsize: 16pt
    footer: "Simulating a market with ABM"
    slide-number: "c"
execute:
  echo: true
---

## The question

:::: {.columns}

::: {.column width="50%"}

Can sellers and buyers show up on an agreed price, without anyone steering?

We will use a toy model with a handful of buyers and sellers, each knowing just one number: the price it's willing to accept or pay. No auctioneer, nobody solving equations. Every round, an agent nudges its expected price up or down depending on whether it managed to trade or not.

I built it with tidyABM, an R package for agent-based models that I wrote.

:::

::: {.column width="50%"}

tidyABM leans on tidyverse habits, and the way you spell out a model owes a lot to NetLogo that focus on reasability. There are 3 main parts:

- setting up the agents, with `abm_setup()`
- defining what happens each period, with `abm_go()`
- running the simulation, with `abm_run()`

The idea comes straight from this Primer video.

[Primer — bottom-up market simulation](https://youtu.be/PNtKXWNKGN8?si=KQebB8FgdtKe-Kl5)

```{r}
pak::pak("rayhanalirachman/tidyABM")
library(tidyABM)
library(tidyverse)
```

:::

::::

## Defining the agents

:::: {.columns}

::: {.column width="50%"}

Lets start small with one seller, two buyers.

Each agent carries four things.

- `limit_price` is the worst price the agent will accept. WTA for a seller, WTP for a buyer.
- `expected_price` is its current guess at a fair price. This is the one that moves each round.
- `transacted` records whether it traded this round.
- `deal_price` records the agreed price, if there was one.

:::

::: {.column width="50%"}

Setup comes down to two questions: how many agents, and what does each one hold. We specify the value of each variables (columns in tidy) directly.

```{r}
market <- abm_setup(
  agents = list(
    seller = abm_agents(
      n = 1,
      limit_price = 20,
      expected_price = 30,
      transacted = FALSE,
      deal_price = NA_real_
    ),
    buyer = abm_agents(
      n = 2,
      limit_price = c(55, 60),
      expected_price = c(40, 45),
      transacted = FALSE,
      deal_price = NA_real_
    )))
```

:::

::::

## What happens each period (tick)

:::: {.columns}

::: {.column width="30%"}
Then we define what happen at each period with `abm_go`. It reads top to bottom, almost like a story. Reset, match, trade and learn.
:::

::: {.column width="70%"}

```{r}
go <- abm_go(
  # 1. Reset the transaction and price
  abm_rules(transacted ~ FALSE, deal_price ~ NA_real_),

  # 2. Match each buyer with a seller at random
  abm_match(pair = "opposite_group", by = .group),

  # 3. Trade if the two sides overlap
  abm_rules(
    # A trade occurs only if the buyer expects at least what the seller expects
    transacted ~ case_when(
      .group == "seller" & partner_expected_price >= expected_price ~ TRUE,
      .group == "buyer"  & expected_price >= partner_expected_price ~ TRUE,
      TRUE ~ FALSE),
    # The deal price is the midpoint of both sides' expectations
    deal_price ~ if_else(
      (.group == "seller" & partner_expected_price >= expected_price) |
      (.group == "buyer"  & expected_price >= partner_expected_price),
      (expected_price + partner_expected_price) / 2, NA_real_)),

  # 4. Learn: each agent nudges its expectation based on the outcome
  abm_rules(
    expected_price ~ case_when(
      .group == "seller" &  transacted ~ expected_price + 1,
      .group == "seller" & !transacted ~ pmax(expected_price - 1, limit_price),
      .group == "buyer"  &  transacted ~ expected_price - 1,
      .group == "buyer"  & !transacted ~ pmin(expected_price + 1, limit_price),
      TRUE ~ expected_price),
    .scope = "population")
)
```

:::

::::

## Running the simulation

`abm_run()` takes three things: the setup, the per-tick logic, and how many ticks to run.

```{r}
result <- abm_run(market, go, ticks = 200, seed = 123) # seed for reproducibility
```

What comes back is a plain tibble, so dplyr works on it right away.

```{r}
result |> slice_head(n = 5)
result |> slice_tail(n = 5)
```

The columns are the ones from the setup, the variable (properties) each agent hold, nothing more.

## Analysing the simulation

:::: {.columns}

::: {.column width="40%"}

From here it's just ggplot. Say we want to watch the deal price move over time.

One seller and two buyers means somebody always goes home empty-handed. That buyer bumps its expectation up next round, so the price drifts upward.

```{r}
result |>
  filter(.group == "buyer") |>
  count(transacted)
```

::: {.callout-note}
More buyers than sellers, and the price climbs.
:::

:::

::: {.column width="60%"}

```{r}
#| echo: false

result |>
  filter(.group == "buyer") |>
  ggplot(aes(x = tick, y = deal_price)) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.6) +
  geom_point(shape = 21, fill = NA, color = "black", alpha = 0.5, stroke = 0.8, size = 1.5) +
  labs(
    title = "Deal price across simulation ticks",
    x = "Simulation tick",
    y = "Deal price") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"))
```

:::

::::

## What if there are 2 sellers and 1 buyer?

:::: {.columns}

::: {.column width="40%"}

Flip the ratio. The model is two blocks, so I swap the setup and leave `go` alone.

```{r}
market2 <- abm_setup(
  agents = list(
    seller = abm_agents(
      n = 2,
      limit_price = c(20, 25),
      expected_price = c(30, 36),
      transacted = FALSE,
      deal_price = NA_real_),
    buyer = abm_agents(
      n = 1,
      limit_price = 60,
      expected_price = 45,
      transacted = FALSE,
      deal_price = NA_real_)))
```

Run it again.

```{r}
result2 <- abm_run(market2, go, ticks = 200, seed = 123)
```

:::

::: {.column width="60%"}

Same plotting code as before, hidden this time.

Now the sellers are the ones competing. The price falls.

```{r}
#| echo: false

result2 |>
  filter(.group == "buyer") |>
  ggplot(aes(x = tick, y = deal_price)) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.6) +
  geom_point(shape = 21, fill = NA, color = "black", alpha = 0.5, stroke = 0.8, size = 1.5) +
  labs(
    title = "Deal price across simulation ticks",
    x = "Simulation tick",
    y = "Deal price") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"))
```

:::

::::

## Many buyers and sellers, balanced

:::: {.columns}

::: {.column width="50%"}

Now scale it up. Fifty on each side. Rather than type out a hundred prices, I draw them from a log-normal.

```{r}
market3 <- abm_setup(
  agents = list(
    seller = abm_agents(
      n = 50,
      limit_price = ~rlnorm(n, log(30), 0.5),
      expected_price = ~rlnorm(n, log(30), 0.7),
      transacted = FALSE,
      deal_price = NA_real_),
    buyer = abm_agents(
      n = 50,
      limit_price = ~rlnorm(n, log(30), 0.5),
      expected_price = ~rlnorm(n, log(30), 0.7),
      transacted = FALSE,
      deal_price = NA_real_)))
```

Run it again.

```{r}
result3 <- abm_run(market3, go, ticks = 200, seed = 5)
```

:::

::: {.column width="50%"}

This plots the mean deal price per tick, with a standard-error ribbon.

The first few ticks are spread so much (look at the shaded area). Then the spread tightens and prices settle around one number. Nobody coordinated that. It's the market mechanism.

```{r}
#| echo: false

result3 |>
  filter(.group == "buyer", !is.na(deal_price)) |>
  dplyr::summarize(
    mean = mean(deal_price),
    se = sd(deal_price) / sqrt(n()),
    .by = tick
  ) |>
  ggplot(aes(x = tick, y = mean)) +
    geom_ribbon(aes(ymin = mean - se, ymax = mean + se), fill = "red", alpha = 0.15) +
    geom_line(color = "red", linewidth = 0.6) +
    labs(
    title = "Mean deal price across simulation ticks",
    x = "Simulation tick",
    y = "Mean deal price") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"))
```

:::

::::

## What I take from it

:::: {.columns}

::: {.column width="55%"}

Small rules, run enough times, add up to something that characterizes market mechanisms.

- Change which side outnumbers the other and the price moves the way the textbook says: up with more buyers, down with more sellers.
- With fifty agents a side, the scattered deal prices pull together into one clearing price. No auctioneer, no equation, nobody in charge of it.

Keeping setup and rules as separate blocks also paid off. Every new scenario was a one-block swap, not a rewrite.

:::

::: {.column width="45%"}

tidyABM is a tidy-friendly ABM package for R, its still experimental but can be checked here

[github.com/rayhanalirachman/tidyABM](https://github.com/rayhanalirachman/tidyABM)


The original market model is lifted from this [Primer video](https://youtu.be/PNtKXWNKGN8?si=KQebB8FgdtKe-Kl5).

Issues and pull requests welcome.

:::

::::
