---
title: "Simulating Market with Agent-Based Model"
author: Rayhan Ali
format : 
  revealjs:
    fontsize: 18pt
    footer: "simulating market with ABM"
    include-in-header:
      text: |
        <style>
        .reveal h2 {
          border-bottom: 2px solid #051C2C;
          padding-bottom: 0.15em;
          margin-bottom: 0.5em;
        }
        </style>
execute:
  echo: true
---

## Setting up
:::: {.columns}

::: {.column width="50%"}
Hello, this is a simulation of a market with buyer and seller within agent-based model that utilise my own package, tidyABM

tidyABM is a ABM package in R that utilise tidy principle. It use similar grammar decision as netlogo, it has 3 main components:

- setting up the agents
- what happen at each period
- running the simulation

It based on primer video about that exact thing that you can see here:

[Youtube](https://youtu.be/PNtKXWNKGN8?si=KQebB8FgdtKe-Kl5)
:::

::: {.column width="50%"}
First we need to install and load the necessary package, we will use tidyverse and tidyABM

```{r}
pak::pak("rayhanalirachman/tidyABM")
library(tidyABM)
library(tidyverse)
```

:::

::::

## Defining the agent

:::: {.columns}

::: {.column width="50%"}

To start, we will use a pretty simple example, with 1 seller and 2 buyer

Each agent will have: 

- Price limit they are willing to accept or pay (i.e. WTA for seller and WTP for buyer)
- Expected price to simulate what the agent expect the price should be, this will be adaptive depending on wether they succesfully trade or not
- And 2 other variable, transacted and deal_price which the information wether the trade is succesfull or not be stored there, and the agreed price

:::

::: {.column width="50%"}

Setting it up is pretty straightforward, how many agent there is (n) and  what variable will each agent needed to have

in this case, we will have 1 seller and 1 buyer

```{r}
market <- abm_setup(
  agents = list(
    seller = abm_agents(
      n = 1,
      limit_price = 20,
      expected_price = 30,
      transacted = FALSE,
      deal_price = NA_real_
    ),
    buyer = abm_agents(
      n = 2,
      limit_price = c(55, 60),
      expected_price = c(40,45),
      transacted = FALSE,
      deal_price = NA_real_
    )
  )
)
```

:::

::::



## Setting up what happen at each period (tick)

:::: {.columns}

::: {.column width="30%"}
To set what happen at each tick, we declare what happen in a descending order, much like a story
:::

::: {.column width="70%"}

```{r}
go <- abm_go(
  # First we set at the beginning that there is no transaction
  abm_rules(transacted ~ FALSE, deal_price ~ NA_real_), 

  # Then we match the buyer and seller randomly
  abm_match(pair = "opposite_group", by = .group),

  # Then simulating the trade
  abm_rules(
    # Transaction only occur if:                                 
    transacted ~ case_when(
      .group == "seller" & partner_expected_price >= expected_price ~ TRUE,
      .group == "buyer"  & expected_price >= partner_expected_price ~ TRUE,
      TRUE ~ FALSE),
    # And the deal price is the middle price of both parties expected price
    deal_price ~ if_else(
      (.group == "seller" & partner_expected_price >= expected_price) |
      (.group == "buyer"  & expected_price >= partner_expected_price),
      (expected_price + partner_expected_price) / 2, NA_real_)),

  # Each agent then update their expected price, given wether they 
  # succesfully trade or not
  abm_rules(                            
    expected_price ~ case_when(
      .group == "seller" &  transacted ~ expected_price + 1,
      .group == "seller" & !transacted ~ pmax(expected_price - 1, limit_price),
      .group == "buyer"  &  transacted ~ expected_price - 1,
      .group == "buyer"  & !transacted ~ pmin(expected_price + 1, limit_price),
      TRUE ~ expected_price),
    .scope = "population")
)
```

:::

::::

## Then run the simulation
To run the simulation, we use abm_run() function with inputting the setup (in this case market), what happen at each tick (go) and how many periods will the model run.
```{r}
result <- abm_run(market, go, ticks = 200, seed = 123) # Using seed for reproducibility
```

The results is in a tibble, which can easily analysed using dplyr. 

```{r}
result |> slice_head(n = 5)
result |> slice_tail(n = 5)
```

As you can see, the column match with what we describe at the setup

## Analysing the simulation

:::: {.columns}

::: {.column width="40%"}

We can analyse the data with ggplot. For example, we want to see how will the agreed price evolve overtime. 

With only 1 seller and 2 buyer, the buyer expected price will always increase, as there will always be 1 buyer that doesnt transact (which in turn raised its expected price).

```{r}
result |> 
  filter(.group == "buyer") |> 
  count(transacted)
```

Thus, price increase.


:::

::: {.column width="60%"}

```{r}
result |> 
  filter(.group == "buyer") |>
  ggplot(aes(x = tick, y = deal_price)) +
  geom_smooth(method = "loess", se = FALSE, color = "red", size = 0.5) +
  geom_point(shape = 21, fill = NA, color = "black", alpha = 0.5, stroke = 0.8, size = 1.5) +
  labs(
    title = "Deal Price Across Simulation Ticks",
    x = "Simulation Tick",
    y = "Deal Price",) +
  theme_minimal() +
  theme(plot.title.position = "plot", 
    plot.title = element_text(face = "bold", size = 13))
```

:::

::: {.callout-note}
This also show when there is more demand (buyer) than supply (seller), price will increase.
:::

::::

## What happens if 2 seller and 1 buyer?

:::: {.columns}

::: {.column width="40%"}

lets now see what happen when there is more seller, as the package is using 2 main blocks, we can just swap the block that need to be changed.

```{r}
market2 <- abm_setup(
  agents = list(
    seller = abm_agents(
      n = 2,
      limit_price = c(20, 25),
      expected_price = c(30, 36),
      transacted = FALSE,
      deal_price = NA_real_),
    buyer = abm_agents(
      n = 1,
      limit_price = 60,
      expected_price = 45,
      transacted = FALSE,
      deal_price = NA_real_)))
```

Then we run it again.

```{r}
result2 <- abm_run(market2, go, ticks = 200, seed = 123) 
```

:::

::: {.column width="60%"}

The plotting code is similar to before, so i will not show it again.

As we would expect, while there is more seller (supply) then buyer (demand), the price will decrease.

```{r}
#| echo = FALSE

result2 |> 
  filter(.group == "buyer") |>
  ggplot(aes(x = tick, y = deal_price)) +
  geom_smooth(method = "loess", se = FALSE, color = "red", size = 0.5) +
  geom_point(shape = 21, fill = NA, color = "black", alpha = 0.5, stroke = 0.8, size = 1.5) +
  labs(
    title = "Deal Price Across Simulation Ticks",
    x = "Simulation Tick",
    y = "Deal Price",) +
  theme_minimal() +
  theme(plot.title.position = "plot", 
    plot.title = element_text(face = "bold", size = 13))
```

:::

::::



## Equal amount many buyer and seller with many agents

:::: {.columns}

::: {.column width="50%"}

Lets up the game now, lets use 50 buyer and 50 seller. But rather than we specify the limit_price and expected_price one by one, we will use a function to specify it

```{r}
market3 <- abm_setup(
  agents = list(
    seller = abm_agents(
      n = 50,
      limit_price = ~rlnorm(n, log(30), 0.5),
      expected_price = ~rlnorm(n, log(30), 0.7),
      transacted = FALSE,
      deal_price = NA_real_),
    buyer = abm_agents(
      n = 50,
      limit_price = ~rlnorm(n, log(30), 0.5),
      expected_price = ~rlnorm(n, log(30), 0.7),
      transacted = FALSE,
      deal_price = NA_real_)))
```

Then we run it again.

```{r}
result3 <- abm_run(market3, go, ticks = 200, seed = 5) 
```

:::

::: {.column width="50%"}

We will only show the mean deal_price and the standard error to show variability. 

As we can see, there are huge variation in the early tick, but overtime, it almost merge to a single price. This is the price mechanism.

```{r}
#| echo = FALSE

result3 |>
  filter(.group == "buyer", !is.na(deal_price)) |>
  dplyr::summarize(
    mean = mean(deal_price),
    se = sd(deal_price) / sqrt(n()),
    .by = tick
  ) |> 
  ggplot(aes(x = tick, y = mean)) +
    geom_ribbon(aes(ymin = mean - se, ymax = mean + se), fill = "red", alpha = 0.15) +
    geom_line(color = "red", linewidth = 0.5) +
    labs(
    title = "Deal Price Across Simulation Ticks",
    x = "Simulation Tick",
    y = "Deal Price") +
    theme_minimal() +
    theme(
      plot.title.position = "plot", 
      plot.title = element_text(face = "bold", size = 13))
```

:::

::::




