---
title: "Solow Growth Model"
format: html
---

This is a short demonstration on simulating basic Solow model within R. The basic solow model posits that the output of an economy ($Y$) is determined by 2 variables, capital ($K$) and labor ($L$), given the following function:
$$Y = K^{\alpha}L^{\beta}$$

where $\alpha$ is the exponent of capital and $\beta$ is the exponent of labor, it is also usually assumed a constant return to scale, where $\alpha + \beta = 1$

The stock of this model is $K$ and $L$, and how the stock evolve (change of stock) is determined by this equation:

$\dot{L} = n \cdot L$ - where n is the population growth

$\dot{K} = s \cdot Y - \delta K$ - where $s$ is the savings rate, and $\delta$ is the depreciation rate of $K$. In this simple model, $s \cdot Y$ is also the investment ($I$), so we assume all that is not consumed become $I$

First we install the necessary package, in this example we focus on using tidyverse package
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(plotly)
```

Then we setup the necessary parameter and initial value
```{r}
# Parameters
alpha <- 1/3   #The exponent of capital
beta <- 2/3    #The exponent of Labor
s <- 0.3       #Savings rate
n <- 0.02      #Pop. Growth
delta <- 0.05  #Depreciation rate

# Initial value
K0 <- 100
L0 <- 100
```

## Building the model
To make it step by step, we can build something like this, where we calculate all the value at t=0. We can divide it into 4 major parts, 

- Stock Variable
- Flow Variable
- The change of stock (due to the flow)
- Ratios (this is optional)
```{r}
state0 <- tibble(
  t = 0, 

  # Stock variable
  L = L0, 
  K = K0, 

  # Flow variable
  Y = (K0^alpha)*(L0^beta),
  I = s*Y, 
  C = (1-s)*Y,

  # Change of stock
  L_dot = n*L0, 
  K_dot = s*Y - delta*K0,

  # Ratios
  k = K0/L0, 
  y = Y/L0
)
```

This create a dataframe of each our value at t=0, given the level of stock is $L0$ and $K0$
```{r}
glimpse(state0)
```

Then we can iterate it again and again until our specified time (Tmax) by creating a function and acummulate it.
```{r}
step_function <- function(state, t) {
  # Calculating this period's stock, which is = previous stock + change of stock (from t-1) 
  L <- state$L + state$L_dot
  K <- state$K + state$K_dot

  # This period's flow, computed from THIS period's stock
  Y <- (K^alpha)*(L^beta)
  I <- s*Y
  C <- (1-s)*Y

  # Change of stock, computed from THIS period's stock (this leads to t+1) 
  L_dot <- n*L
  K_dot <- s*Y - delta*K

  # Ratiosatio
  k <- K/L
  y <- Y/L

  # Then we collect it into a dataframe which feed to the next cycle
  tibble(t, L, K, L_dot, K_dot, Y, I, C, k, y)
}

# Using accumulate function is simplifying things a lot, or we can also use the for loop function
path_list <- accumulate(1:200, step_function, .init = state0)
```

Then we can plot it to see how the model evolves

First is the plot of $Y$ to $t$, as we can see the output is growing indefinitely as the $K$ and $L$ also grow.
```{r}
path_list |> 
    plot_ly(x = ~t, y = ~Y, type = "scatter", mode = "lines")
```

But we can also see the growth rate of Y, as we see, the growth rate of $Y$ settles at around 0.02, which we call this a steady state (basically just means the value is not changing anymore)

```{r}
path_list |> 
    plot_ly(x = ~t, y = ~(Y/ lag(Y))-1, type = "scatter", mode = "lines")
```

Or usually the more helpful is not $Y$ per se, but the output to labor ratio ($y = \frac{Y}{L}$)
.
```{r}
path_list |> 
    plot_ly(x = ~t, y = ~y, type = "scatter", mode = "lines")
```

However, in the solow model, we are more interested on capital to labor ratio ($k$), which eventually arrived at a steady state too

```{r}
path_list |> 
    plot_ly(x = ~t, y = ~y, type = "scatter", mode = "lines")
```

## Simulating a shocks
We can also simulate a shock here, like for example, there is a sudden drop of savings rate (due to crisis etc)
```{r}
s <- 0.3 # this is the initial savings rate

path_1 <- accumulate(1:200, step_function, .init = state0) #running t 1:200

s <- 0.2   # <- new parameter value after t = 200

path_2 <- accumulate(201:400, step_function, .init = path_1[nrow(path_1), ]) #running after change in parameters

path_list2 <- bind_rows(path_1, path_2  |> slice(-1)) #Combining the 2 dataframe, before and after shock
```

We can plot this again, as we can see the model achieve new steady state of $y$ 
```{r}
path_list2 |> 
    plot_ly(x = ~t, y = ~y, type = "scatter", mode = "lines")
```

And also new steady state for $k$
```{r}
path_list2 |> 
    plot_ly(x = ~t, y = ~k, type = "scatter", mode = "lines")
```

## Solow Diagram
We can also make the solow diagram here, i will not explain the derivatives here but basically its a diagram explaining the rate of $k$ that produce steady state.

As the change of $k$ is
$$\dot{k} = s \cdot k^\alpha - (n+\delta) \cdot k $$

For steady state, $\dot{k} = 0$ thus steady state achieve where

$$\underbrace{s \cdot k^\alpha}_{\text{Actual Investment}} = \underbrace{(n+\delta) \cdot k}_{\text{Break-Even Investment}}$$

We can visualise this
```{r}
s <- 0.2

solow_df <- tibble(
  k = seq(0.01, 20, length.out = 200),
  Actual_investment = s * k^alpha,
  Break_even = (n + delta) * k
)

solow_df |> 
    plot_ly(x = ~k) |> 
    add_trace(y = ~Actual_investment, type = "scatter", mode = "lines", name = "Actual Investment") |> 
    add_trace(y = ~Break_even, type = "scatter",mode = "lines", name = "Break Even Investment") |> 
    layout(
      title =  "Solow-Swan Steady State Diagram",
      yaxis = list(title = "Investment per worker", zeroline = TRUE),
      legend = list(x = 0.05, y = 0.95)
    )
```

We can also see what happens if there is a shock, e.g. an increase in $s$, which results to upward shift to actual investment curve, pushing steady state $k$ higher
```{r}
s <- 0.2
s_2 <- 0.4

solow_df <- tibble(
  k = seq(0.01, 20, length.out = 200),
  Actual_investment = s * k^alpha,
  Actual_investment_2 = s_2 * k^alpha,
  Break_even = (n + delta) * k
)

solow_df |> 
    plot_ly(x = ~k) |> 
    add_trace(y = ~Actual_investment, type = "scatter", mode = "lines", name = "Initial Actual Investment") |> 
    add_trace(y = ~Actual_investment_2, type = "scatter", mode = "lines", name = "New Actual Investment") |> 
    add_trace(y = ~Break_even, type = "scatter",mode = "lines", name = "Break Even Investment") |> 
    layout(
      title =  "Solow-Swan Steady State Diagram",
      yaxis = list(title = "Investment per worker", zeroline = TRUE),
      legend = list(x = 0.05, y = 0.95)
    )
```